Equation-fitting: Table 4, Thornthwaite and Mather (1957)
========================================================

This documents the process of fitting a general equation to Table 4 of Thornthwaite and Mather (1957). Table 4 is entitled "Values of unadjusted daily potential evapotranspiration (mm) for different mean temperatures (degrees C) and I values." This equation fitting exercise is designed as a check of the regression equations coded into the original version of SWB.

The first step is to define the general form of the equation we wish to use to summarize the Thornthwaite-Mather tables. The original version of SWB defines this table with a quartic (4th degree) polynomial:
```{r}
# x=I; y=Temp_degC
gen<-function(x, y,fac0, fac1, fac2, fac3, fac4, fac5) {

  result <- fac0 + fac1*x + fac2*y + fac3*x*y + fac4*x^2 + fac5*y^2
  
  return(result)
}

```

The next code block reads in the data from Thornthwaite and Mather (1957), Table 2. 


```{r}

library(minpack.lm) 

load("D:\\SMWData\\Source_Code\\swb\\share\\Table_summaries\\TM_Table_4\\TM_Table_4.RData")
mycolnames <- colnames(TM_table_4)[2:ncol(TM_table_4)]

TM_4C <- TM_table_4


```

The potential evapotranspiration at various combinations of temperature and "I" values looks like this:


```{r, fig.width=14, fig.height=10}

#filled.contour(x=x, y=y, z=TM4_mat, xlab="THORNTHWAITE-MATHER \"I\" VALUE",
##      ylab="AIR TEMPERATURE, IN DEGREES FAHRENHEIT", col=topo.colors(30),
#      main="POTENTIAL EVAPOTRANSPIRATION, IN INCHES")


```

```{r}

TM4C_yxz <- data.frame(Temp_degC=numeric(), I=numeric(), POTET=numeric() )

mycolnames <- colnames(TM_table_4)[2:ncol(TM_table_4)]

y<- TM_table_4[ ,1]
x<- as.numeric(gsub(pattern="I", replacement="", x=mycolnames) )

temp_df <-data.frame(Temp_degC=y, I=numeric(length(y)), POTET=numeric(length(y)) )

for (i in 2:ncol(TM_4C)) {

  temp_df$I <- x[i-1]
  temp_df$POTET <- TM_4C[ ,i]

  TM4C_yxz <- rbind(TM4C_yxz, temp_df)
}

# sort in ascending order
TM4C_yxz <- TM4C_yxz[order(TM4C_yxz[ ,c(1:2)]), ]
TM4C_yxz <- subset(TM4C_yxz, !is.na(POTET))

TM4C_mat <- TM4C_yxz[ ,c(1,2)]

TM_4C_opt <- nlsLM( POTET ~ gen(I, Temp_degC, fac0, fac1, fac2, fac3, fac4, fac5), 
                  data=TM4C_yxz, start=list(fac0=0., fac1=0.04, fac2=0.02, fac3=0.05, fac4=0.01, fac5=0.01),
                  control=list(maxiter=1024, tol=1.e-13, printEval=TRUE, 
                               minFactor=1.e-8, warnOnly=TRUE))


fac <- as.numeric(TM_4C_opt$m$getPars())

```


```{r, fig.width=15, fig.height=48}

layout(matrix(c(1:32), nrow=16 , ncol=2, byrow = TRUE))

for (airtemp in seq(0,26.5,0.5)) {
  
  I <- seq(25,140,5)
  
  table_ET <- subset(TM4C_yxz, airtemp==Temp_degC)
  predicted_ET <- gen(I, airtemp, fac[0], fac[1], fac[2], fac[3], fac[4], fac[5])
  
  with(table_ET, plot(xlim=c(20,145),ylim=c(0,7), x=I, y=POTET, cex=0.7) )
  lines(x=I, y=predicted_ET, col="red")
  
  
  
}


```
