Adventures with the curve number
========================================================

First we'll define a couple of key curve number-related functions.

```{r CNfunctions, message=FALSE, warning=FALSE, results='hide'}

library(sfsmisc)

Smax <- function(CN) {
  
  value <- (1000 / CN) - 10
  return(value)
  
}

Ia <- function(Smax) {
  
  value <- 0.2 * Smax
  return(value)
  
}
```
Let's calculate Smax and Ia for a range of curve numbers.

```{r cn_calcs, message=FALSE, warning=FALSE, results='hide'}
CN <- seq(1,98,1)

Smax_values <- Smax(CN)
Ia_values <- Ia(Smax(CN))
```

First, we'll plot the maximum storage term against our curve numbers.

```{r fig_Smax, message=FALSE, warning=FALSE, results='hide'}
plot (Smax_values ~ CN, xlab="Curve Number", ylab="Smax (inches)", col="red")
```
The originators of curve number methodology state that Smax has a range of 0 to infinity, so that plot seems to agree with them.

Next, let's plot the initial abstraction term values as a function of the curve number.

```{r fig_Ia, message=FALSE, warning=FALSE, results='hide'}
plot(Ia_values ~ CN, xlab="Curve Number", ylab="Initial Abstraction (inches)",   log="y",axes=FALSE, col="blue", main="Initial abstraction for medium density residential (CN=89)")
aY <- c(0.001,0.01,0.1,1,10.,100.)
axis(2, at=aY, label= aY, las=2)
axis(1)
box()
grid(equilogs=FALSE)
abline(v=89)
my_Smax <- Ia(Smax(89))
abline(h=my_Smax)
text(y=my_Smax +0.1, x=40,paste("Smax =",formatC(my_Smax,digits=2) ) )
box()
```
Now we'll repeat using the curve number for a soil from hydrologic soils group A with cultivated crops (CN=71).

```{r fig_Ia_ag, message=FALSE, warning=FALSE, results='hide'}
plot(Ia_values ~ CN, xlab="Curve Number", ylab="Initial Abstraction (inches)",   log="y",axes=FALSE, col="blue", main="Initial abstraction for cultivated crops, A soils (CN=71)")
aY <- c(0.001,0.01,0.1,1,10.,100.)
axis(2, at=aY, label= aY, las=2)
axis(1)
box()
grid(equilogs=FALSE)
abline(v=71)
my_Smax <- Ia(Smax(71))
abline(h=my_Smax)
text(y=my_Smax +0.1, x=40,paste("Smax =",formatC(my_Smax,digits=2) ) )
box()
```